Deploying and Sharing QSAR Models using Predictive Modeling Markup Language

### 2014-12-02 paper.tex
Updated some text, added margin comments

\documentclass[12pt,letterpaper]{article}
\usepackage{fullpage, times}
\usepackage{hyperref}
\usepackage{ctable}
\begin{document}
\title{Deploying and Sharing QSAR Models using Predictive Modeling Markup Language}
\author{Villu Ruusmann${}^{\dagger}$ and Rajarshi Guha${}^{\ddagger}$\\
${}^{\ddagger}$National Center for Advancing Translational Sciences\\ 9800 Medical Center Drive  Rockville, MD 20850 \\ \\
${}^{\dagger}$Your address }
\date{}
\maketitle
\begin{abstract}
  A nice abstract
\end{abstract}

@@

Quantitative Structure-Activity Relationship (QSAR) models have become
ubiquitous in computational chemistry and cheminformatics, having been
built for a variety of endpoints including biological activities and
physical properties. As shown in Figure \ref{fig:count-qsar} the last
15 years has seen a significant increase in publications developing or
using QSAR models. While there has been much discussion (REFS) on the
prospective utility and validity of predictive QSAR models, a key
bottleneck to their assessment is the fact that they are not easily
obtainable. That is, the bulk of QSAR models are described in a
publication in terms of the descriptors (sometimes made available) and
the modeling method. While the implementation is noted, it is upon the
reader to actually obtain the descriptor values, obtain and set up the
modeling environment and then actually rebuild the model. In other
words, the actual model that was developed by the authors is, usually,
not available to be used directly. In a few cases, the model may be
built in a pipeline tool [REF], which alleviates this problem to some
extent. Furthermore, the model development workflow usually involves
multiple software tools, each one of a specific version. Without
access to the exact same version of these tools, a user cannot be
guaranteed to exactly reproduce the model that the authors
describe. The issue of reproducibility (of workflows and results) has
become increasingly important [REF Greg Landrum] in the computational
arena.

A number of workers have described infrastructure to deploy predictive
models. Some platforms such as Pipeline Pilot provide a easy to use
method to deploy a complete pipeline as a web page. However, this
approach does not allow one to share models with users who do not have
access to the software. Guha [REF] described a web based
infrastructure that allowed one to deploy serialized models developed
in R. This architecture allowed one to distribute the model via a web
page, allowing users to provide new input and retrieve
predictions. But by virtue of using serialized R models, it allowed
the user to directly work with the original model itself. An important
downside was that the model itself did not contain explicit references
to the descriptors that were required as input.

@@

The main question: if you see a QSAR model in the literature, do you
want to analyze its mathematical properties, or do you want to apply
it to your dataset in order to see how it performs?

@@

The ONS approach is to use Wiki pages, which contain embedded R code
and plots. For example, see
\url{http://onschallenge.wikispaces.com/MeltingPointModel002}.

@@

Essentially, the representation of a model training process boils down
to representing the training algorithm. The most complete representation
is in application code such as R, Python, MATLAB etc.
The algorithm "works" by applying a finite list of well-defined 
instructions to the initial state. After applying each instruction the
system reaches another state. When the list of instructions has been
exhausted the system has reached its final ending state.
The representation of states is technically much more challenging than 
the representation of algorithms. A state may consist of different
parts. We are only interested in that part, which relates to the 
mathematical relationship. A state can be preserved by taking a snapshot
of the particular memory area. To make reading and writing more easier,
should be formalized as a data structure.

@@

The QsarDB approach [ref?] attempts to capture the initial state, the 
final ending state and all states inbetween using a collection of files
in open and standard data formats. For example, datasets are represented
in TSV, parameter units in UCUM [ref?], bibliography in BibTeX [ref?], 
mathematical relationships in PMML etc. QsarDB leaves the door open
for extension information is user-defined data formats. For example, a
QsarDB archive may capture the primary model object that corresponds to 
the "main" run, as well as all secondary model objects that correspond to 
validation runs.

@@

A model represents a standalone mathematical relationship. Not very easy
to work with from the end user perspective. A model should be packaged
as a predictive solution, which has the following components:
\begin{itemize}
  \item Data schema. Specifies information about input and output fields.
  Data schema represents the "public interface" for the end user. 
  For example, data schema could communicate that a predictive solution
  consumes a single string-type input field that contains the SMILES 
  representation of a chemical structure, and produces two output fields - 
  a string-type field that contains the classification result (eg. the 
  predicted mechanism of toxic action) and a floating point-type field 
  that contains a class-specific regression result (eg. if the chemical
  structure was classified as a narcotic agent, the predicted pIGC50 
  value using the narcosis QSAR model).
  \item Data pre- and post-processing functions. Typically, the input
  fields that are listed in data schema cannot be directly fed to the 
  underlying matchematical relationship "as is". Could be related to 
  translating input from user representation to internal representation
  (eg. converting SMILES to CDK molecular graph to CDK descriptor value)
  or performing additional transformations on it (eg. normalization, 
  converting from natural scale to logaritmic scale etc.).Similarly,
  could derive secondary output fields based on the primary field. For
  example, calculating the probability of the predicted class or standard
  error of the predicted value.
  \item Data (work)flows. A predictive solution may encapsulate more
  than one model. The passage of data from one model to another happens
  inside the predictive solution. Two most common scenarious is model 
  aggregation, where the final prediction is calculated by applying an
  aggregation function over a list of interim predictions (eg. consensus
  modeling), and model chaining, where the output of earlier models
  becomes the input of subsequent models (eg. classification followed 
  by regression).
\end{itemize}

@@

It is possible to distinguish between three levels:
\begin{itemize}
  \item Data structure. Tied to a specific library and/or algorithm.
  For example, the \textit{randomForest} R library returns the trained
  RF model as a special-purpose data structure. Moreover, the internal
  structure of this data structure varies depending on whether the RF
  model was trained using the "formula interface" or the "matrix interface".
  Such data structures can be easily persisted into files using so-called
  serialization mechanisms. The storing operation (ie. saving the in-memory
  data structure to disk file) is called serialization, and the loading
  operation (ie. reading the disk file data structure to memory) is called
  deserialization. Most language runtimes provide built-in serialization
  capabilities. For example, R language provides methods \textit{saveRDS}
  and \textit{readRDS} that can work with binary and text-based files.
  \item Pseudocode. A similar model can be trained using different libraries
  and/or algorithms. For example, a RF model can also be trained using
  the \textit{party} R library (or in-house R library). The choice of the
  algorithm determines how the search is conducted in the feature space. 
  However, the final ending state of a RF training process is an ensemble
  of decision tree models.
  The goal of the pseudocode representation is to extract and encapsulate
  the primary structure of the model. The pseudocode representation aims 
  to be much more informative for humans (eg. following the computation
  path from beginning to end), but still be directly interpretable by 
  machines. For example, a decision tree model could be represented by
  a tree-like data structure, where non-leaf nodes specify split criteria 
  and leaf nodes specify predicted scores.
  \item Application code. During deployment, the model is executed with
  with new input data. With data structure and pseudocode approaches this 
  is accomplished by model interpretation. The alternative approach is to
  translate either representation to application code, which would allow
  for more optimizations and much speedier execution. 
\end{itemize}

Binary serialization mechanisms are typically much more performant and 
compact than text serialization. The only downside, which is rather minor,
is that binary files are hard to open and handle without special tools.
With text files it is possible to take a peek and possibly even make minor
modifications.

@@

A molecular descriptor is a characteristic that is obtained by applying
a particular algorithm to a particular chemical structure representation.
Typically classified by the complexity of structure representation:
\begin{enumerate}
  \item 1D. Atom counts.
  \item 2D. Atom connectivity.
  \item 3D. Atom locations.
\end{enumerate}

Making descriptor calculation reproducible is difficult because it is
very tightly coupled with the preparation of chemical compounds to a suitable
chemical structure representation. For example, consider a simplest 2D
molecular descriptor that simply counts the number of aromcatic carbon
atoms. The functional definition of aromaticity is still a matter of 
ongoing debate -- classical organic chemistry interpretation (Hückel rule) 
vs. modern theoretical/computational interpretation.

By definition, when the algorithm is applied to an incorrect chemical
structure representation, the results is an incorrect molecular descriptor
value. Cannot reliably say if an user-specified representation is correct
or not by merely "looking" at it. Will need to apply a rigorous verification
procedure. For example, suppose SMILES chemical structure representation.
There exist different variations of SMILES, some of which make aromaticy
transparent by using different atom and bond symbols (ie. lowercase letter "c" 
and/or carbons connected by ":" bonds). Therefore, simple analysis of SMILES
using simple text analysis functions won't suffice, and it has to be parsed 
and reconstructed into more sophisticated chemical graph. Even then it is
possible to run into trouble, because the ring detection and ring aromaticity
assignment can be performed following different algorithms.

When "interfacing" QSAR models with actual end-users then it is necessary
to specify supported input chemical structure representations. Most likely,
actual chemical databases contain chemical compounds in data formats that
cannot be fed \emph{directly} to the molecular descriptor calculation
algorithm. For example, when testing with new chemical compounds then user
input may originate from graphical drawing board (eg. JMol applet [ref?] or 
J(S)ME Java(Script) Molecular Editor applet [ref?]) or be a systematic
name. The desirable property of a QSAR predictive solution would be to
recognize and automatically convert between different chemical structure
representations.

@@

There have been attempts to "simplify" the systematizing and communicating 
algorithms using the ontology approach. The earliest work is the Blue Obelisk
Descriptor Ontology [ref?], which has been now superseded by the Chemical 
Information (CHEMINF) Ontology [ref?]. Unfortunately, not much adoption, 
because the vocabulary of real-life algorithms is way beyond the standardized
vocabularies, and the lack of competing implementations. An ontology that
has been crafted following CDK data structures and concepts is unlikely to
be relevant and useful for other descriptor calculation software. In other
words, if there is an one-to-one correspondence between ontology identifiers
and CDK implementation classes, then it may be a better idea to give those
CDK class names directly, and avoid the intermediary layers of "ontology 
identifiers" altogether.

The futility of formalizing algorithms using the ontology approach is
well demonstrated by the "parametrization problem". The behaviour of most
algorithms can be tweaked by adjusting a set of control variables. For 
example, when performing the analysis of 3D molecular shapes, finding the 
optimum between accuracy and speed (eg. mesh size). A different
parametrization will likely result in different calculated molecular
descriptor values. Cannot have repeatable nor less reproducible science
when the exact parametrization is unavailable. The only way to address
this problem would be to extend the algorithm ontology with this extra
information, but that would limit its relevancy and usefulness even more.

Another source of "variance" for molecular descriptor algorithms are 
programming errors. It is possible that newer versions of the same software 
could yield different results than older versions, because the algorithm
has been slightly changed (or reparameterized) in some point in time.

@@

For this study we developed models in R 3.1 [REF] with descriptors
generated using the CDK [REF] via the rcdk [REF] package. As exemplars
we considered two datasets. XXX Describe the datasets.

@@

>>>>

\documentclass[12pt,letterpaper]{article}
\usepackage{fullpage, times}
\usepackage{hyperref}
\usepackage{ctable}

%% margin notes - comment out before submission
\usepackage{marginnote}
\usepackage[top=2.54cm, bottom=2.54cm, outer=5cm, inner=1.25cm, heightrounded, marginparwidth=4cm, marginparsep=0.25in]{geometry}
\renewcommand*{\marginfont}{\color{red}\small}

\bibliographystyle{jcics}

\begin{document}
\title{Deploying and Sharing QSAR Models using Predictive Modeling Markup Language}
\author{Villu Ruusmann${}^{\dagger}$ and Rajarshi Guha${}^{\ddagger}$\\
${}^{\ddagger}$National Center for Advancing Translational Sciences\\ 9800 Medical Center Drive  Rockville, MD 20850 \\ \\
${}^{\dagger}$Your address }
\date{}
\maketitle
\begin{abstract}
  A nice abstract
\end{abstract}

@@

Quantitative Structure-Activity Relationship (QSAR) models have become
ubiquitous in computational chemistry and cheminformatics, having been
built for a variety of endpoints including biological activities and
physical properties. As shown in Figure \ref{fig:count-qsar} the last
15 years has seen a significant increase in publications developing or
using QSAR models. While there has been much discussion
\cite{Cherkasov:2014yq,Cramer2011Inevitable,Tropsha:2003aa} on the
prospective utility and validity of predictive QSAR models, a key
bottleneck to their assessment is the fact that they are not easily
obtainable. That is, the bulk of QSAR models are described in a
publication in terms of the descriptors (sometimes made available) and
the modeling method. While the implementation is noted, it is upon the
reader to actually obtain the descriptor values, obtain and set up the
modeling environment and then actually rebuild the model. In other
words, the actual model that was developed by the authors is, usually,
not available to be used directly. In a few cases, the model may be
built in a pipeline tool \cite{KNIME:2008aa}, which alleviates this
problem to some extent. Furthermore, the model development workflow
usually involves multiple software tools, each one of a specific
version. Without access to the exact same version of these tools, a
user cannot be guaranteed to exactly reproduce the model that the
authors describe. The issue of reproducibility (of workflows and
results) has become increasingly important \cite{Landrum:2012hl,
  Stodden:2013gd,Lane:2003fe} in the computational arena.

A number of workers have described infrastructure to deploy predictive
models. Some platforms such as Pipeline Pilot provide a easy to use
method to deploy a complete pipeline as a web page. However, this
approach does not allow one to share models with users who do not have
access to the software. Guha \cite{Guha:2007af} described a web based
infrastructure that allowed one to deploy serialized models developed
in R. This architecture allowed one to distribute the model via a web
page, allowing users to provide new input and retrieve
predictions. While this solution allowed the deployment of models,
they were not actually reproducible, since the serialized model did
not actually reference the input descriptors.

@@

When dealing with a published QSAR model, one must choose whether to
analyze its mathematical properties, or apply it to a dataset in order
to see how it performs?

@@

The ONS approach is to use Wiki pages, which contain embedded R code
and plots. For example, see
\url{http://onschallenge.wikispaces.com/MeltingPointModel002}. More
recently, websites such as
\url{https://rpubs.com/}{https://rpubs.com/} support direct web
publishing from ones local development environment.

@@

\marginnote{This paragraph reads a bit too abstract. Can it be stated
  more concisely?}Essentially, the representation of a model training process boils down
to representing the training algorithm. The most complete representation
is in application code such as R, Python, MATLAB etc.
The algorithm "works" by applying a finite list of well-defined 
instructions to the initial state. After applying each instruction the
system reaches another state. When the list of instructions has been
exhausted the system has reached its final ending state.
The representation of states is technically much more challenging than 
the representation of algorithms. A state may consist of different
parts. We are only interested in that part, which relates to the 
mathematical relationship. A state can be preserved by taking a snapshot
of the particular memory area. To make reading and writing more easier,
should be formalized as a data structure.

@@

The QsarDB approach \cite{Aruoja:2014bq,Ruusmann:2014uk} attempts to
capture the initial state, the final ending state and all states
inbetween using a collection of files in open and standard data
formats. For example, datasets are represented in TSV, parameter units
in UCUM [ref?], bibliography in BibTeX [ref?], mathematical
relationships in PMML etc. QsarDB leaves the door open for extension
information is user-defined data formats. For example, a QsarDB
archive may capture the primary model object that corresponds to the
"main" run, as well as all secondary model objects that correspond to
validation runs.

@@

A predictive model represents a standalone mathematical
relationship. The model itself, in terms of its mathematical form, is
not particularly easy to work with from the end user perspective. For
actual usage, a model should be packaged as a predictive solution,
which has the following components:
\begin{itemize}
  \item Data schema. Specifies information about input and output fields.
  Data schema represents the "public interface" for the end user. 
  For example, data schema could communicate that a predictive solution
  consumes a single string-type input field that contains the SMILES 
  representation of a chemical structure, and produces two output fields - 
  a string-type field that contains the classification result (eg. the 
  predicted mechanism of toxic action) and a floating point-type field 
  that contains a class-specific regression result (eg. if the chemical
  structure was classified as a narcotic agent, the predicted pIGC50 
  value using the narcosis QSAR model).
  \item Data pre- and post-processing functions. Typically, the input
  fields that are listed in data schema cannot be directly fed to the 
  underlying matchematical relationship "as is". Could be related to 
  translating input from user representation to internal representation
  (eg. converting SMILES to CDK molecular graph to CDK descriptor value)
  or performing additional transformations on it (eg. normalization, 
  converting from natural scale to logaritmic scale etc.).Similarly,
  could derive secondary output fields based on the primary field. For
  example, calculating the probability of the predicted class or standard
  error of the predicted value.
  \item Data (work)flows. A predictive solution may encapsulate more
  than one model. The passage of data from one model to another happens
  inside the predictive solution. Two most common scenarious is model 
  aggregation, where the final prediction is calculated by applying an
  aggregation function over a list of interim predictions (eg. consensus
  modeling), and model chaining, where the output of earlier models
  becomes the input of subsequent models (eg. classification followed 
  by regression).
\end{itemize}

@@

It is possible to distinguish between three levels:
\begin{itemize}
  \item Data structure. Tied to a specific library and/or algorithm.
  For example, the \textit{randomForest} R library returns the trained
  RF model as a special-purpose data structure. Moreover, the internal
  structure of this data structure varies depending on whether the RF
  model was trained using the "formula interface" or the "matrix interface".
  Such data structures can be easily persisted into files using so-called
  serialization mechanisms. The storing operation (ie. saving the in-memory
  data structure to disk file) is called serialization, and the loading
  operation (ie. reading the disk file data structure to memory) is called
  deserialization. Most language runtimes provide built-in serialization
  capabilities. For example, R language provides methods \textit{saveRDS}
  and \textit{readRDS} that can work with binary and text-based files.
  \item Pseudocode. A similar model can be trained using different libraries
  and/or algorithms. For example, a RF model can also be trained using
  the \textit{party} R library (or in-house R library). The choice of the
  algorithm determines how the search is conducted in the feature space. 
  However, the final ending state of a RF training process is an ensemble
  of decision tree models.
  The goal of the pseudocode representation is to extract and encapsulate
  the primary structure of the model. The pseudocode representation aims 
  to be much more informative for humans (eg. following the computation
  path from beginning to end), but still be directly interpretable by 
  machines. For example, a decision tree model could be represented by
  a tree-like data structure, where non-leaf nodes specify split criteria 
  and leaf nodes specify predicted scores.
  \item Application code.\marginnote{Its not clear what this level
      means. Are you making a distinction between 'interpretation' and
      'compilation'? It seems data structures or pseudocode both end
      up as application code at one point}During deployment, the model is executed with
  with new input data. With data structure and pseudocode approaches this 
  is accomplished by model interpretation. The alternative approach is to
  translate either representation to application code, which would allow
  for more optimizations and much speedier execution. 
\end{itemize}

Binary serialization mechanisms are usually more performant and 
compact compared to text based methods. The primary disadvantage is
that they cannot be manually inspected and require special tools to
parse and view them.

@@

A molecular descriptor is a characteristic that is obtained by applying
a particular algorithm to a particular chemical structure representation.
Typically classified by the complexity of the structure representation:
\begin{enumerate}
  \item 1D - Atom and bond counts and other constitutional features
  \item 2D - Atom-atom connectivity
  \item 3D - Conformation dependent, 3D atom coordinates
\end{enumerate}

Descriptor calculations can be difficult to reproduce. This is
primarily driven by differences in the chemistry models implemented in
different toolkits and chemical software platforms. An example is the
count of aromatic carbons - which is dependent on the aromaticity
model (of which there are several) employed by the implementation. As
a result, while descriptors calculated using a given toolkit are
reproducible, this is only guaranteed within that \emph{specific
  version} of the toolkit. The same algorithm may give different
results using another cheminformatics toolkit.

\marginnote{This paragraph seems to digress a little and doesn't
  really focus on the issue at hand}By definition, when the algorithm
is applied to an incorrect chemical structure representation, the
results is an incorrect molecular descriptor value. Cannot reliably
say if an user-specified representation is correct or not by merely
"looking" at it. Will need to apply a rigorous verification
procedure. For example, suppose SMILES chemical structure
representation.  There exist different variations of SMILES, some of
which make aromaticy transparent by using different atom and bond
symbols (ie. lowercase letter "c" and/or carbons connected by ":"
bonds). Therefore, simple analysis of SMILES using simple text
analysis functions won't suffice, and it has to be parsed and
reconstructed into more sophisticated chemical graph. Even then it is
possible to run into trouble, because the ring detection and ring
aromaticity assignment can be performed following different
algorithms.

\marginnote{Is it really the case that the structure representation is
  the issue? Shouldn't any decent standardization tool transform multiple
  input forms to a single canonical form?}Due to these issues,
providing a ``complete'' QSAR models to an end-user requires that we
specify the supported input chemical structure representations. Most
likely, actual chemical databases contain chemical compounds in data
formats that cannot be fed \emph{directly} to the molecular descriptor
calculation algorithm. For example, when testing with new chemical
compounds then user input may originate from graphical drawing board
(eg. JMol applet [ref?] or J(S)ME Java(Script) Molecular Editor applet
[ref?]) or be a systematic name. The desirable property of a QSAR
predictive solution would be to recognize and automatically convert
between different chemical structure representations.

@@

There have been attempts to "simplify" the systematizing and
communicating algorithms using the ontology approach. The earliest
work is the Blue Obelisk Descriptor Ontology \cite{Guha:2006ac}, which
has been now superseded by the Chemical Information (CHEMINF) Ontology
\cite{Hastings:2011rf}.

\marginnote{Probably not useful to speak negatively (at least in this
  article). Maybe drop this paragraph or just simplify
  it}Unfortunately, not much adoption, because the vocabulary of
real-life algorithms is way beyond the standardized vocabularies, and
the lack of competing implementations. An ontology that has been
crafted following CDK data structures and concepts is unlikely to be
relevant and useful for other descriptor calculation software. In
other words, if there is an one-to-one correspondence between ontology
identifiers and CDK implementation classes, then it may be a better
idea to give those CDK class names directly, and avoid the
intermediary layers of "ontology identifiers" altogether.

The challenges of formalizing algorithm descriptions using ontologies
is well demonstrated by the ``parametrization problem''.  That is, the
behaviour of most algorithms can be tweaked by adjusting a set of
control variables or parameters. For example, one approach to 3D
analysis of molecular shape requires one to specify a mesh
size. Different values represent trade-offs between speed and accuracy
and will lead to different descriptor values. Thus a full
specification of an algorithm must also include the details of the
parametrization. Encoding this level of detail in an ontology may not
be appropriate.

\marginnote{This is always a possibility - not sure whether its worth
  mentioning}Another source of "variance" for molecular descriptor
algorithms are programming errors. It is possible that newer versions
of the same software could yield different results than older
versions, because the algorithm has been slightly changed (or
reparameterized) in some point in time.

@@

For this study we developed models in R 3.1 [REF] with descriptors
generated using the CDK \cite{Steinbeck:2003bh} via the rcdk
\cite{Guha:2007aa} package. As exemplars we considered two
datasets. XXX Describe the datasets.

@@

\bibliography{paper}
